Retrofitting Effect Handlers onto OCaml

### 2020-10-29 retro-concurrency.tex
Section 3

\usepackage{balance}
\usepackage{url}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xspace}
\usepackage{cleveref}
\usepackage{multirow}
\usepackage{microtype}
\usepackage{soul,xcolor}
\usepackage{mathpartir}
\usepackage{amsmath,amsthm}
\usepackage[utf8]{inputenc}
\input{unicode-preamble.tex}
\usepackage{mathtools}
\usepackage{tikz}
\usetikzlibrary{calc,decorations.pathmorphing,shapes}
\usepackage{stmaryrd}

@@

\lstset{
			language=caml,
      basicstyle=\ttfamily\footnotesize,
      flexiblecolumns=false,
			tabsize=2,
			escapechar={<@}{@>},
      %basewidth={0.5em,0.45em},
      %aboveskip={3pt},
      %belowskip={3pt},
			breaklines = true,
      breakatwhitespace = true,
      keywordstyle=\color{Bittersweet}\bfseries,
      commentstyle=\color{blue}\itshape,
      stringstyle=\color{MidnightBlue},
			keywords=[2]{},
			keywordstyle=[2]\color{byzantine}\bfseries,
			keywords=[3]{effect,perform,continue,discontinue},
			keywordstyle=[3]\color{caribbeangreen}\bfseries,
			keywords=[4]{call,return,external},
			keywordstyle=[4]\color{Bittersweet}\bfseries,
			classoffset=1,
			upquote=true,
			keywordstyle=\color{byzantine}\bfseries,
			classoffset=0,
			mathescape=true,
			numberstyle=\tiny\color{gray},
			numbersep=5pt
}

@@

The computation |gsearch(1/0)| returns positive infinity (|Inf P|). We can
extend the calculator modularly with the ability to read user input using the
|Read| effect defined earlier as follows:

@@

OCaml, however, does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations is
to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Koka}. If the compiler were to use CPS, then reifying
continuations in the source language is relatively straight-forward. But OCaml
does not use CPS IR, and changing the OCaml compiler to utilise CPS IR would be
an enormous undertaking. CPS IR would also affect the performance profile of
OCaml applications due to increased memory allocations as continuation closures
get allocated on the heap~\cite{Folklore}. Moreover, with CPS, stack unwinding
is meaningless and hence, we would lose compatibility with tools that inspect
the program stack. Hence, we choose not to CPS translation.

The alternative is to use the call stack for implementing continuation.
Farvardin et al~\cite{Folklore} show that there is a spectrum of implementation
strategies for continuation with different trade-offs. We motivate our
implementation strategy based on the following ideal requirements:

\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility} A well behaved OCaml program does not
		break under OCaml extended with effect handlers. OCaml code that does not
		use effect handlers will pay negligible performance cost.
	\item \textbf{Tool compatibility} OCaml with effect handlers produce
		well-formed backtraces, and remains compatible with program analysis tools
		such as debuggers and profilers that inspect the stack using DWARF unwind
		tables.
	\item \textbf{Effect handler efficiency} In order to support
		highly-concurrent applications, the program must accommodate millions of
		continuations in the heap. Installing effect handlers, capturing and
		resuming continuations must be fast.
	\item \textbf{Effect handler expressivity} While making continuations
		resumable at-most once (\emph{affine}) is a standard way to make
		continuation capture fast, we must still support use cases where the
		continuations are resumed more than once.
\end{enumerate}

The need to host millions of active continuations in the heap rules out the use
of large contiguous stack space for continuations as in C. Instead, we resort
to using small initial stacks and growing the stacks on demand. As a result,
OCaml functions, irrespective of whether they use effect handlers, need to
perform stack overflow checks and external C functions, which do not have stack
overflow checks, must be performed on a separate stack. Additionally, we must
generate DWARF stack unwind tables for stacks that may be non-contiguous and
contain frames from different languages. In this work, we develop the compiler
and runtime support for implementing efficient effect handlers for OCaml.

@@

The search for an expressive effect system that guarantees that all the effects
performed in the program are handled (\emph{effect safety}) in the presence of
advanced features such as polymorphism, modularity and generativity is an
active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
Hillerstrom20}. We do not focus on this question in this paper, and our
implementation of effect handlers in OCaml does not guarantee effect safety.
For example, running |perform (Read stdin)| without an enclosing handler raises
exception |Unhandled|. We leave the question of effect safety for future work.

@@

The rest of the paper is organized as follows. We describe the details of the
OCaml program stack in the next section, and highlight the features that the
effect handler implementation must retain in order to provide backwards
compatibility. In \S\ref{sec:handlers}, we motivate the design choice that we
have made for effect handlers in Multicore OCaml with the help of examples.
\S\ref{sec:semantics} presents the formal operational semantics for Multicore
OCaml effect handlers. \S\ref{sec:impl} discusses the compiler and the runtime
system support for implementing effect handlers. \S\ref{sec:eval} presents
extensive performance evaluation of effect handlers against the goals set in
\S\ref{sec:req}. \S\ref{sec:related} and \S\ref{sec:conc} present the related
work and the conclusions.

@@

\begin{figure*}
\begin{minipage}{0.59\linewidth}
	\begin{minipage}{\linewidth}
		\begin{lstlisting}
external ocaml_to_c : unit -> int = "ocaml_to_c"
exception E1
exception E2

@@

let _ = assert (omain () = 42)
		\end{lstlisting}
    \subcaption{\texttt{meander.ml}}
    \label{code:meander_ml}
	\end{minipage}
	\begin{minipage}{\linewidth}
		\begin{lstlisting}[language=c]
#include <caml/mlvalues.h>
#include <caml/callback.h>

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. On the |amd64| backend, a register each is
pinned for the \emph{allocation pointer} into the minor heap used for bump
pointer allocation, and the |Caml_state|, a table of global variables used by
the runtime. These registers are chosen to be C callee saved registers. This
makes external calls extremely fast in OCaml. If the external function does not
touch the OCaml heap, then the external functions can be called directly, and
no bookkeeping is necessary. For external functions that touch the OCaml heap,
the cached allocation pointer is saved to a field in the |Caml_state|, and
loaded back on return. Similarly, callbacks into OCaml are also cheap.
Callbacks into OCaml involves loading the arguments in the right registers and
calling the OCaml function. OCaml callbacks are relatively common as
finalisation functions are executed as OCaml callbacks.

@@

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Hence, installing an exception handler simply pushes the program counter (pc)
of the handler and the current exception pointer (a field in |Caml_state|). The
exception pointer is also updated to be the current stack pointer (sp). This
creates a linked-list of exception handler frames on the stack as shown in
Figure~\ref{fig:stock_stack}. The saved exception pointer is restored when an
exception is handled or the handled computation returns normally with a value.

@@

\kc{gdb backtrace}

\section{Handlers by example}
\label{sec:handlers}

>>>>

\usepackage{balance}
%\usepackage{url}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xspace}
\usepackage{cleveref}
\usepackage{multirow}
\usepackage{microtype}
\usepackage{soul,xcolor}
\usepackage{mathpartir}
\usepackage{amsmath,amsthm}
\usepackage[utf8]{inputenc}
\input{unicode-preamble.tex}
\usepackage{mathtools}
\usepackage{tikz}
\usetikzlibrary{calc,decorations.pathmorphing,shapes}
\usepackage{stmaryrd}

@@

\lstset{
			language=caml,
      basicstyle=\ttfamily\footnotesize,
      flexiblecolumns=false,
			tabsize=2,
			escapechar={<@}{@>},
      %basewidth={0.5em,0.45em},
      %aboveskip={3pt},
      %belowskip={3pt},
			breaklines = true,
      breakatwhitespace = true,
      keywordstyle=\color{Bittersweet}\bfseries,
      commentstyle=\color{blue}\itshape,
      stringstyle=\color{MidnightBlue},
			keywords=[2]{},
			keywordstyle=[2]\color{byzantine}\bfseries,
			keywords=[3]{effect,perform,continue,discontinue,clone_continuation},
			keywordstyle=[3]\color{caribbeangreen}\bfseries,
			keywords=[4]{call,return,external},
			keywordstyle=[4]\color{Bittersweet}\bfseries,
			classoffset=1,
			upquote=true,
			keywordstyle=\color{byzantine}\bfseries,
			classoffset=0,
			mathescape=true,
			numberstyle=\tiny\color{gray},
			numbersep=5pt
}

@@

The computation |gsearch(1/0)| returns positive infinity (|Inf P|).

\subsection{Asynchronous I/O}
\label{sec:aio}

We can extend the calculator modularly with the ability to read user input
using the |Read| effect defined earlier as follows:

@@

OCaml, however, does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Koka}. If the compiler were to use CPS, then reifying
continuations in the source language is relatively straight-forward. But OCaml
does not use CPS IR, and changing the OCaml compiler to utilise CPS IR would be
an enormous undertaking. CPS IR would also affect the performance profile of
OCaml applications due to increased memory allocations as continuation closures
get allocated on the heap~\cite{Folklore}. Moreover, with CPS, stack unwinding
is meaningless and hence, we would lose compatibility with tools that inspect
the program stack. Hence, we choose not to CPS translation. The alternative is
to use the call stack for implementing continuation. Farvardin et
al~\cite{Folklore} show that there is a spectrum of implementation strategies
for continuation with different trade-offs.

The search for an expressive effect system that guarantees that all the effects
performed in the program are handled (\emph{effect safety}) in the presence of
advanced features such as polymorphism, modularity and generativity is an
active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
Hillerstrom20}. We do not focus on this question in this paper, and our
implementation of effect handlers in OCaml does not guarantee effect safety.
For example, running |perform (Read stdin)| without an enclosing handler raises
exception |Unhandled| at the point of |perform|. We leave the question of
effect safety for future work.

\subsection{Requirements}
\label{sec:requirements}

We motivate our implementation strategy based on the following ideal
requirements:

\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility} A well behaved OCaml program does not
		break under OCaml extended with effect handlers. OCaml code that does not
		use effect handlers will pay negligible performance cost.
	\item \textbf{Tool compatibility} OCaml with effect handlers produce
		well-formed backtraces, and remains compatible with program analysis tools
		such as debuggers and profilers that inspect the stack using DWARF unwind
		tables.
	\item \textbf{Effect handler efficiency} In order to support
		highly-concurrent applications, the program must accommodate millions of
		continuations in the heap. Installing effect handlers, capturing and
		resuming continuations must be fast.
	\item \textbf{Effect handler expressivity} While making continuations
		resumable at-most once (\emph{affine}) is a standard way to make
		continuation capture fast, we must still support use cases where the
		continuations are resumed more than once.
	\item \textbf{Forwards compatibility} While this work does not present an
		effect system, the design must be compatible with a future effect system
		for OCaml. As a cornerstone of modularity, we also want blocking I/O code
		to \emph{transparently} be made asynchronous with the help of effect
		handlers.
\end{enumerate}

The need to host millions of active continuations in the heap rules out the use
of large contiguous stack space as in C for continuations. Instead, we resort
to using small initial stacks and growing the stacks on demand. As a result,
OCaml functions, irrespective of whether they use effect handlers, need to
perform stack overflow checks and external C functions, which do not have stack
overflow checks, must be performed on a separate stack. Additionally, we must
generate DWARF stack unwind tables for stacks that may be non-contiguous and
contain frames from different languages. In this work, we develop the compiler
and runtime support for implementing efficient effect handlers for OCaml.

@@

@@

The rest of the paper is organized as follows. We describe the details of the
OCaml program stack in the next section, and highlight the features that the
effect handler implementation must retain in order to provide backwards
compatibility. In \S\ref{sec:refine}, we refine the design further
focussing on the issues that come up when integrating effect handlers into a
mainstream language. \S\ref{sec:semantics} presents the formal operational
semantics for Multicore OCaml effect handlers. \S\ref{sec:impl} discusses the
compiler and the runtime system support for implementing effect handlers.
\S\ref{sec:eval} presents extensive performance evaluation of effect handlers
against the goals set in \S\ref{sec:req}. \S\ref{sec:related} and
\S\ref{sec:conc} present the related work and the conclusions.

@@

\begin{figure*}
\begin{minipage}{0.59\linewidth}
	\begin{minipage}{\linewidth}
		\begin{lstlisting}[numbers=left]
external ocaml_to_c : unit -> int = "ocaml_to_c"
exception E1
exception E2

@@

let _ = assert (omain () = 42)
		\end{lstlisting}
    \subcaption{\texttt{meander.ml}}
    \label{code:meander_ml}
	\end{minipage}
	\begin{minipage}{\linewidth}
		\begin{lstlisting}[language=c,numbers=left]
#include <caml/mlvalues.h>
#include <caml/callback.h>

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. On the |x86-64| backend, a register each is
pinned for the \emph{allocation pointer} into the minor heap used for bump
pointer allocation, and the |Caml_state|, a table of global variables used by
the runtime. These registers are chosen to be C callee saved registers. This
makes external calls extremely fast in OCaml. If the external function does not
allocate in the OCaml heap, then the external functions can be called directly,
and no bookkeeping is necessary. For external functions which allocate in the
OCaml heap, the cached allocation pointer is saved to a field in the
|Caml_state| before the external call, and loaded back on return. Similarly,
callbacks into OCaml are also cheap. Callbacks into OCaml involves loading the
arguments in the right registers and calling the OCaml function. OCaml
callbacks are relatively common as the garbage collector (GC) finalisation
functions are executed as OCaml callbacks.

@@

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Hence, installing an exception handler simply pushes the program counter (pc)
of the handler and the current exception pointer (|exn_ptr| -- a field in
|Caml_state|). The exception pointer is also updated to be the current stack
pointer (|sp|). This creates a linked-list of exception handler frames on the
stack as shown in Figure~\ref{fig:stock_stack}. Raising an exception simply
sets |sp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps to the |pc| of the
handler.

@@

\subsection{DWARF unwind tables}

OCaml also generates precise DWARF unwind information for OCaml, thanks to
which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
and~\ref{code:meander_c}, one could set a break point in |gdb| at
|caml_raise_exn| to get the following backtrace.

\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
% gdb ./meander.exe
(gdb) break caml_raise_exn
Breakpoint 1 at 0x3e5dc
(gdb) r
Starting program: meander.exe
Breakpoint 1, 0x925dc in caml_raise_exn ()
(gdb) bt
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_start_program ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander_.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:14
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_start_program ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}

\noindent which corresponds to the stack layout in
Figure~\ref{fig:stock_stack}.

The same backtrace can also be obtained by using \emph{frame pointers} instead
of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
they are not enabled by default. OCaml stack tends to be deep with small frames
due to the pervasive use of recursive functions, not all of which is
tail-recursive. Hence, the addition of frame pointers can significantly
increase the size of the
stack~\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
Moreover, not using frame pointers saves two instructions in the function
prologue and epilogue, and makes an extra register (|rbp| on x86) available.

In summary, the design of effect handlers in OCaml must accommodate these
desirable properties of OCaml with respect to the program stack.

\section{Refining the design}
\label{sec:refine}

In this section, we refine the effect handler design for the OCaml use case.
The primary motivation for adding effect handlers to OCaml is to support
non-local control flow mechanism to implement asynchronous I/O in direct style,
generators and enable task parallelism. These use cases do not resume the
continuations more than once. Hence, our continuations are one-shot, and
resuming the continuation more than once raises |Invalid_argument _| exception.
It is well-known that one-shot continuations can be implemented
efficiently~\cite{Bruggeman96RepresentingControl}.

While OCaml permits throwing exceptions across C frames, we do not allow
effects to propagate across C frames as the C frames would become part of the
captured continuation. Managing C frames as part of the continuation is a
complex endeavour~\cite{Daan17Aplas}, and we find that the complexity budget
outweighs the relatively fewer mechanisms enabled by this addition in our
setting.

\subsection{Discontinuing a continuation}

The interaction of non-local control flow with systems programming is quite
subtle~\cite{}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:

\begin{lstlisting}
let rec copy ic oc =
  try
    let l = input_line ic in
    output_string oc (l ^ "\n");
    copy ic oc
  with _ -> (close_in ic; close_out oc)
\end{lstlisting}

|input_line| raises |End_of_file| exception on reaching the end of input.
Although not apparent, the code is written in a defensive style to handle other
exceptional cases such as the channels being closed externally. Both
|input_line| and |output_string| raise |Sys_error| exception if the channel is
closed. This function cleanly closes the channels on exceptional cases;
|close_*| functions do nothing if the channel is already closed.

One of our goals(\S\ref{sec:requirements}) is to make this code transparently
asynchronous. We can define effects for performing the I/O operations:
\begin{lstlisting}
effect In_line : in_channel -> string
effect Out_str : out_channel -> string -> unit
\end{lstlisting}
\noindent and wrap them up in functions with the same signature as the one from
the standard library.
\begin{lstlisting}
let input_line ic = perform (In_line ic)
let output_string oc = perform (Out_str oc)
\end{lstlisting}

We can then use a top-level I/O handler similar to the introductory example
(\S\ref{sec:aio}) to discharge the I/O operations asynchronously. This handles
value return cases, but what about the exceptional cases |End_of_file| and
|Sys_error|? We introduce a |discontinue| primitive to resume a continuation by
raising an exception. In this example, on reaching the end of file, we would
discontinue the continuation of the |copy| function with |discontinue k End_of_file|,
which raises the exception at |input_line|, and the open channels
will be closed.

Program that use resources such as channels are usually written defensively
with the assumption that calling a function will return \emph{exactly once},
either normally or exceptionally. Since effect handlers in Multicore OCaml do
not ensure that all the effects are handled, if the function performs an effect
with no matching handler, then the function \emph{will not return at all}. To
remedy this, when such an effect bubbles up to the top-level, we discontinue
the continuation with |Unhandled| exception so that the exception handlers may
run and clean up the resources.

\subsection{Multi-shot continuations}

Though our primary goal is to enable non-local control flow mechanisms, effect
handlers have been gaining increasing attention for Bayesian machine learning
and probabilisitic programming~\cite{PyroJMLR,MooreEdward2Google}, which resume
the continuations more than once for performing inference over the
probabilistic program. To enable these use cases, we provide a
|clone_continuation| primitive that makes a copy of the given continuation.
Care must be taken when using multi-shot continuations as they do not mix well
with code that uses resources. With multi-shot continuations, a function
\emph{may return more than once}. Consider the code below:

\begin{lstlisting}
let foo m = ...;perform E; ...
let bar () =
  let m = malloc 64 in
	begin try foo m with _ -> () end;
	free m
let _ = match bar () with
| () -> ()
| effect E k ->
    let k' = clone_continuation k in
		continue k (); continue k' ()
\end{lstlisting}

The function |bar| expects |foo| to return exactly once, ignoring any exceptions. However,
due to continuation cloning, the function returns twice, leading to double-free
error.
